Income Inequality is revealed in Torontonian Apartment Maintenance

### 2022-02-04 outputs/paper/paper.Rmd
Added text to the readme file.
Made several changes and modifications to these files.

---
title: "Income Inequality is revealed in Torontonian Apartment Maintenance"
subtitle: "My subtitle if needed"
author: 
  - Olaedo Okpareke
  - Another author
thanks: "Code and data are available at: https://github.com/OlaedoOkpareke/starter_folder."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Housing maintenance beneifts residents as it ensures proper functioning of ammenities, maintains structural integrity, and adds value to homes. However, previous research has found that priority in maintenance for public housing apartments and lower income apartments is little compared to their more privileged counterparts. We use data from Open data Toronto to further analyze this by using plots and tables to break down apartment scores and other attributes by Ward and apartment type, using R (R Core Team 2021) and other packages. We find that apartments in higher income areas and privately owned apartments tend to have better evaluation scores than those in lower income wards and public housing respectively. This data shows the inequality in standards for housing in Toronto, as well as the need for better government accountability and housing justice. Improving the gap in aparment quality helps improve the life of poor and marginalized people."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---
BOXPLOTS TALK ABOYT MEDIANS< NOT MEANS. PUT THE TABLE FOR MEAN
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
install.packages("opendatatoronto")
library(opendatatoronto)
```

@@

```{r}
apartment_data = list_package_resources("4ef82789-e038-44ef-a478-a8f3590c3eb1") %>%
  filter(name=="Apartment Building Evaluation") %>%
  get_resource()

apartment_data = na_if(apartment_data, 'N/A')

rent_data <- apartment_data[complete.cases(apartment_data), ]

# Try  to filter by year evaluated and see if it gets what I want
rent_clean<- rent_data %>% 
  dplyr::select(Score = SCORE,
                Ward = WARDNAME,
                Year_built = YEAR_BUILT,
                Type= PROPERTY_TYPE,
                Security = SECURITY
  )

rent_clean$Score = strtoi(rent_clean$Score)
rent_clean$Year_built = as.double(rent_clean$Year_built)
rent_clean$Security = as.double(rent_clean$Security)
```

```{r}
rent_clean %>%
  type_convert(
  rent_clean,
  col_types = cols(
    Score = col_integer(),
    Ward = col_character(),
    Year_built = col_double(),
    Type = col_character(),
    Security = col_double()
  ),
  na = c("", "N/A"),
  guess_integer = FALSE
)
```

We are interested in the relationships between property Type and Ward on Score and security, as well as the trend of Score over time. It shows a measure of income inequality in housing, and has possible implications for public policy. A preview of the data, (Table \@ref(tab:1)) using kableexrra(Zhu 2021) is shown below.
```{r 1, fig.cap = "Table of the first few rows of the data"}
x = head(rent_clean) %>%
  knitr::kable(col.names = c("Score", "Ward", "Year built","Type", "Security"), caption = "Toronto Apartment Evaluation Data")
x
```

(Table \@ref(tab:2)) shows the the summary statistics for the buildings. We can see from the table that the average building has good standards and security, but is a bit old. The average apartment has a better score than the median apartment, indicating that there are some apartments that are far above the average. However the mean security is lower the median, indicating some very low security apartments. Policy is needed to improve security for these apartments, and the observation that some scores are far higher than most are more thoroughly scrutinized later in the report.

```{r 2, fig.cap = "Table with summaries of the Score and Security for all apartments"}
rent_clean %>% 
  summarise(mean=mean(rent_clean$Score),median=median(rent_clean$Score), mean1 = mean(rent_clean$Security),median2=median(rent_clean$Security)) %>%
  knitr::kable(col.names = c("Mean Score", "Median Score", "Mean Security","Median Security"), caption = "Table of Summaries for Apartment Maintenance Scores")
```

(Figure @\ref(fig: 1)), using ggplot2 (Wickham 2016), shows the average maintenance scores by ward. Cross referencing the income and demographics of each ward using (City of Toronto 2022) , we see that the higher income wards such as Toronto Saint Paul's and Toronto-Danforth (155,470 and 101,323 average yearly income respectively 72% white and 67% white respectively as of 2016) tend to have both higher means and medians than lower income wards, such as Davenport and Humber River-Black Creek (80,807 and 65,458 average yearly income as of 2016). An exception seems to be Scarborough North which has a low yearly average income and high racial minority population but a high average apartment maintenance score. However note that the range is very small, implying that there are very few samples, so they may not be representative of the true state of apartments in this ward. Although there are several outlier Scores for each ward, in general apartment housing is better in richer wards than in poorer wards. High income Wards also tend to have fewer racial minorities, with Toronto saint Paul and Toronto-Danforth having a 72% and 67% non-racial minority population respectively.
```{r 1, fig.cap = "Boxplot of median apartment maintenance scores by ward", echo = TRUE}
rent_clean %>%
  ggplot(aes(x = Ward,y = Score,group = Ward)) + 
  theme_minimal() +
  geom_boxplot() +
  coord_flip()+
  labs(title = "Wealthier wards have better scores than poorer Wards",
       x = "Ward",
       y = "Apartment Score")
```

```{r ,fig.cap = "Barplot of Mean Apartment maintenance scores by ward}
rent_clean %>%
  ggplot(aes(x = Ward,y = Score)) + 
  theme_minimal() +
  geom_bar()
  coord_flip()+
  labs(title = "Wealthier wards tend to have better apartment maintenance than poorer",
       x = "Ward",
       y = "Average Apartment Score")

```

(Figure @\ref(fig:2)) shows the mean and median apartment scores by type.  The results show that private buildings have the best average maintenance quality, and those from the Toronto housing Corporation have the lowest average. The spread of quality seems to be equal for all three housing types, so even accounting for variation private housing is consistently better than their public counterparts. This shows a disparity in priority for apartment maintenance and upkeep, as well as general housing inequality as those in social housing owned by the TCHC also deserve high average maintenance but do not, and reveals that the government are not on top of taking care of social housing. However it should be noted that (non TCHC owned) social housing maintenance is not very far behind private housing, showing a just upkeep of housing. Nevertheless according to (), people in social housing often have to take maintenance and upkeep into their own hands. 
```{r 2,fig.cap = "Boxplot of median apartment scores by Apartment Type", echo = TRUE }
rent_clean %>%
  ggplot(aes(x = Type,y = Score)) + 
  theme_minimal() +
  geom_boxplot() +
  labs(title = "Private Housing has better scores than social and TCHC housing ",
       x = "Apartment Type",
       y = "Apartment Score")
```
(Table \@ref(tab:3))  shows the differences in security between each housing type. We see that overall the average secuirty scores are good for all types of apartment. However the median shows that both TCHC owned and non TCHC owned social buildings have less stellar security than that of the private buildings, so their average scores were driven up by  a few very good apartments. This means that we can expect that a large number of social housing (which houses mainly low income populations) do not have the best security, which is counter-intuitive as poorer areas where this housing is often located are often in higher need of security due to correlation with crime rates. (cite poor housing live in crime areas). This could be alleviated with multi-income neighbourhoods, which places many different types of housing that can accomodate different income groups, but NIMBY-ism has slowed the progress of these neighbourhoods.
```{r}
rent_clean %>%
  group_by(Type) %>%# We want to know the occupancy by month
  summarize(Mean = mean(Security),Median = median(Security)) %>%
  knitr::kable()
```

(Figure @\ref(fig:3)) shows the general trend of apartment maintenance scores for all types according to the year they were built. Even though for all three types newer buildings have better maintenance Scores than their older counterparts, private housing still gets consistently higher scores than the other two types for almost all the years. Most private apartments were built between the 1960s to 1970s while social housing was built between the 1980s to 2000s, so even the newer age of the social housing buildings is not enough to protect against lower rates of maintenance. This shows either the persistence in lack of government and public maintenance or persistent bias in scoring, which both tie into income inequality. It should be noted that the trend for private apartments may be heavily influenced by the points at the far right end, and that many private apartments also have low scores. We also removed an outlier, 1910 as it affected the accuracy of the trend for private apartments. 
```{r 3,fig.cap = "Scatterplots and regression line of trend of apartment scores by type over 1950 - 2021", echo = FALSE }
rent_clean %>%
  filter(Year_built >=1950) %>% # The only value before 1950 (1910, is an outlier)
  ggplot(aes(x = Year_built, y = Score, color = Type)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F)+
  geom_abline()+
  theme_minimal() +
  labs(title = "Privately owned apartment scores have ongiong advantage over social counterparts",
       x = "Year",
       y = "Score",
       color = "Type") +
  scale_color_brewer(palette = "Dark2")
```

(Figure @\ref(fig:4)) shows that the wealthier wards have a larger percentage of their apartments scoring above 70 compared to lower income wards. Some of the poorer wards such as Humber River Black creek or Davenport do not have very highly rated apartments at all. The racial minority population in these two wards as of 2016 was 74% and 43% respectively (City of Toronto 2022).In 2016 the income for these wards was 65,548 and 80,807 dollars respectively while in the richer wards like Toronto Saint Paul and Etobicoke Centre it was 155,470 and 128,448 (City of Toronto 2022). Many other wards follow the same trend, with the wards with higher income having larger percentages of their apartments scoring above 70, with some lower income wards even having some percentage of their apartments have a score less than 50. This is another way to see the inequality in housing standards, and communicates that housing maintenance disproportionally caters to higher income areas. This is possibly because lower income people pay less rent, are less likely to protest (August 2016), or less general funding compared to previous decades (Tsenkova 2021).  note tsenkova states that less funding to build new housing

```{r 4,fig.cap = "Percentage of apartments in each ward with that have a score in a certain group", echo = FALSE }
rent_clean %>%
  mutate(Maintenance_Score = 
           case_when(
             Score <= 50 ~ "Less than 50",
             Score <= 60 ~ "50 to 60",
             Score <= 70 ~ "60 to 70",
             Score <= 80 ~ "70 to 80",
             Score <= 100 ~ "80 or more",
             TRUE ~ "None"
             )
  ) %>%
  ggplot(data=.)+
  theme_minimal()+
  coord_flip()+
  geom_bar(mapping = aes(x = Ward, fill = Maintenance_Score),position="fill")+
  geom_text(aes(label=Maintenance_Score), vjust=1.6, color="white", size=3.5)+
  scale_y_continuous(labels = scales::percent)
# Add Dark2 here 
```

@@

## Income
It is important to note that not all living in private apartments are wealthy. In fact, there are many private apartments inhabited by low income people so this disparity in scores between private and social housing may be due to reasons other than income inequality. As stated earlier, less governmental priority to social apartments may be a major reason (Tsenkova 2021).  The average income for each ward was an important variable as it would put other information into context, as it determines the priority in government spending not just in apartments but in general as lower income areas get less spending but wealthier income are often full of funding in housing, surroundings, and other outside amenities that could have gone into the scoring bias. Unfortunately this variable was not included in the dataset, so information on income had to come from a less up to date source.

@@

\appendix

# Appendix {-}

# Additional details

>>>>

---
title: "Income Inequality is revealed in Torontonian Apartment Maintenance"
author: 
  - Olaedo Okpareke
thanks: "Code and data are available at: https://github.com/OlaedoOkpareke/starter_folder."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Housing maintenance beneifts residents as it ensures proper functioning of ammenities, maintains structural integrity, and adds value to homes. However, previous research has found that priority in maintenance for public housing apartments and lower income apartments is little compared to their more privileged counterparts. We use data from Open data Toronto to further analyze this by using plots and tables to break down apartment scores and other attributes by Ward and apartment type, using R (R Core Team 2021) and other packages. We find that apartments in higher income areas and privately owned apartments tend to have better evaluation scores than those in lower income wards and public housing respectively. This data shows the inequality in standards for housing in Toronto, as well as the need for better government accountability and housing justice. Improving the gap in aparment quality helps improve the life of poor and marginalized people."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---
BOXPLOTS TALK ABOYT MEDIANS< NOT MEANS. PUT THE TABLE FOR MEAN
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# install.packages("opendatatoronto")
library(opendatatoronto)
```

@@

```{r}
rent_clean = read_csv(here::here("inputs/data/rent_clean.csv"))
```

We are interested in the relationships between property Type and Ward on Score and security, as well as the trend of Score over time. It shows a measure of income inequality in housing, and has possible implications for public policy. A preview of the data, (Table \@ref(tab:tableone)) and figure 1 using kableexrra(Zhu 2021) and ggplot2 (Wickham...) is shown below.

```{r tableone, fig.cap = "Table of the first few rows of the data"}
x = head(rent_clean) %>%
  knitr::kable(col.names = c("Score", "Ward", "Year built","Type", "Security"), caption = "Toronto Apartment Evaluation Data")
x
```
```{r figone, fig.cap = "Table with summaries of the Score and Security for all apartments"}
ggplot(rent_clean, aes(x=Score)) +
  geom_histogram(fill="black")+
  theme_minimal()
```

(Figure \@ref(fig:figone)) shows the the summary statistics for the buildings. We see that the data seems to be unimodal and normally distributed. There is an outlier which skews the data. We can see from the table that the average building has good standards and security, but is a bit old. The average apartment has a better score than the median apartment, indicating that there are some apartments that are far above the average. However the mean security is lower the median, indicating some very low security apartments. Policy is needed to improve security for these apartments, and the observation that some scores are far higher than most are more thoroughly scrutinized later in the report.

(Figure \@ref(fig:figtwo)), using ggplot2 (Wickham 2016), shows the average maintenance scores by ward. Cross referencing the income and demographics of each ward using (City of Toronto 2022) , we see that the higher income wards such as Toronto Saint Paul's and Toronto-Danforth (155,470 and 101,323 average yearly income respectively 72% white and 67% white respectively as of 2016) tend to have both higher means and medians than lower income wards, such as Davenport and Humber River-Black Creek (80,807 and 65,458 average yearly income as of 2016). An exception seems to be Scarborough North which has a low yearly average income and high racial minority population but a high average apartment maintenance score. However note that the range is very small, implying that there are very few samples, so they may not be representative of the true state of apartments in this ward. Although there are several outlier Scores for each ward, in general apartment housing is better in richer wards than in poorer wards. High income Wards also tend to have fewer racial minorities, with Toronto saint Paul and Toronto-Danforth having a 72% and 67% non-racial minority population respectively.

```{r figtwo, fig.cap = "Boxplot of median apartment maintenance scores by ward", echo = TRUE}

rent_clean %>%
  ggplot(aes(x = Ward,y = Score,group = Ward)) + 
  theme_minimal() +
  geom_boxplot() +
  coord_flip()+
  labs(title = "Wealthier wards have better scores than poorer Wards",
       x = "Ward",
       y = "Apartment Score")
```

(Figure \@ref(fig:figthree)) shows the mean and median apartment scores by type.  The results show that private buildings have the best average maintenance quality, and those from the Toronto housing Corporation have the lowest average. The spread of quality seems to be equal for all three housing types, so even accounting for variation private housing is consistently better than their public counterparts. This shows a disparity in priority for apartment maintenance and upkeep, as well as general housing inequality as those in social housing owned by the TCHC also deserve high average maintenance but do not, and reveals that the government are not on top of taking care of social housing. However it should be noted that (non TCHC owned) social housing maintenance is not very far behind private housing, showing a just upkeep of housing. Nevertheless according to (), people in social housing often have to take maintenance and upkeep into their own hands.

```{r figthree, fig.cap = "Boxplot of median apartment scores by Apartment Type", echo = TRUE }
rent_clean %>%
  ggplot(aes(x = Type,y = Score)) + 
  theme_minimal() +
  geom_boxplot() +
  labs(title = "Private Housing has better scores than social and TCHC housing ",
       x = "Apartment Type",
       y = "Apartment Score")
```

(Table \@ref(tab:tabtwo))  shows the differences in security between each housing type. We see that overall the average secuirty scores are good for all types of apartment. However the median shows that both TCHC owned and non TCHC owned social buildings have less stellar security than that of the private buildings, so their average scores were driven up by  a few very good apartments. This means that we can expect that a large number of social housing (which houses mainly low income populations) do not have the best security, which is counter-intuitive as poorer areas where this housing is often located are often in higher need of security due to correlation with crime rates. (cite poor housing live in crime areas). This could be alleviated with multi-income neighbourhoods, which places many different types of housing that can accomodate different income groups, but NIMBY-ism has slowed the progress of these neighbourhoods.
```{r tabtwo}
rent_clean %>%
  group_by(Type) %>%# 
  summarize(Mean = mean(Security),Median = median(Security)) %>%
  knitr::kable(caption = "Table of Mean and Median security for each group") 
```

(Figure @\ref(fig:figfour)) shows the general trend of apartment maintenance scores for all types according to the year they were built. Even though for all three types newer buildings have better maintenance Scores than their older counterparts, private housing still gets consistently higher scores than the other two types for almost all the years. Most private apartments were built between the 1960s to 1970s while social housing was built between the 1980s to 2000s, so even the newer age of the social housing buildings is not enough to protect against lower rates of maintenance. This shows either the persistence in lack of government and public maintenance or persistent bias in scoring, which both tie into income inequality. It should be noted that the trend for private apartments may be heavily influenced by the points at the far right end, and that many private apartments also have low scores. We also removed an outlier, 1910 as it affected the accuracy of the trend for private apartments. 
```{r figfour,fig.cap = "Scatterplots and regression line of trend of apartment scores by type over 1950 - 2021", echo = FALSE }
rent_clean %>%
  filter(Year_built >=1950) %>% # The only value before 1950 (1910, is an outlier)
  ggplot(aes(x = Year_built, y = Score, color = Type)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F)+
  geom_abline()+
  theme_minimal() +
  labs(title = "Privately owned apartment scores have ongiong advantage over social counterparts",
       x = "Year",
       y = "Score",
       color = "Type") 
```

(Figure @\ref(fig:figfive)) shows that the wealthier wards have a larger percentage of their apartments scoring above 70 compared to lower income wards. Some of the poorer wards such as Humber River Black creek or Davenport do not have very highly rated apartments at all. The racial minority population in these two wards as of 2016 was 74% and 43% respectively (City of Toronto 2022).In 2016 the income for these wards was 65,548 and 80,807 dollars respectively while in the richer wards like Toronto Saint Paul and Etobicoke Centre it was 155,470 and 128,448 (City of Toronto 2022). Many other wards follow the same trend, with the wards with higher income having larger percentages of their apartments scoring above 70, with some lower income wards even having some percentage of their apartments have a score less than 50. This is another way to see the inequality in housing standards, and communicates that housing maintenance disproportionally caters to higher income areas. This is possibly because lower income people pay less rent, are less likely to protest (August 2016), or less general funding compared to previous decades (Tsenkova 2021).  note tsenkova states that less funding to build new housing

```{r figfive,fig.cap = "Percentage of apartments in each ward with that have a score in a certain group", echo = FALSE }
rent_clean %>%
  mutate(Maintenance_Score = 
           case_when(
             Score <= 50 ~ "Less than 50",
             Score <= 60 ~ "50 to 60",
             Score <= 70 ~ "60 to 70",
             Score <= 80 ~ "70 to 80",
             Score <= 100 ~ "80 or more",
             TRUE ~ "None"
             )
  ) %>%
  ggplot(data=.)+
  theme_minimal()+
  coord_flip()+
  geom_bar(mapping = aes(x = Ward, fill = Maintenance_Score),position="fill")+
  scale_y_continuous(labels = scales::percent)
# Add Dark2 here 
```

@@

## Income
It is important to note that not all living in private apartments are wealthy. In fact, there are many private apartments inhabited by low income people so this disparity in scores between private and social housing may be due to reasons other than income inequality. As stated earlier, less governmental priority to social apartments may be a major reason (Tsenkova 2021).  The average income for each ward was an important variable as it would put other information into context, as it determines the priority in government spending not just in apartments but in general as lower income areas get less spending but wealthier income are often full of funding in housing, surroundings, and other outside amenities that could have gone into the scoring bias. Unfortunately this variable was not included in the data set, so information on income had to come from a less up to date source.

@@
